---
title: "Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`stomachr` provides a tidy pipeline for downloading, reading, cleaning, and preparing ICES stomach content data for various diet analysis. The data come from the [ICES stomach content database](https://stomachdata.ices.dk) and follow the new four-table exchange format, joined by `tblUploadID`, `tblHaulID`, and `tblPredatorInformationID`.

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("maxlindmark/stomachr")
```

## Workflow

### Step 1: Download

You can download the data from the portal and go to Step 2. Or you can call `download_stomach()`, which downloads data from the ICES API and writes the four CSVs to a local directory (which you control via `path`). You can filter by `year`, `country`, and `ecoregion` (one of `"Baltic Sea"`, `"Celtic Seas"`, `"Greater North Sea"`).

```{r download}
path <- "data/raw"
download_stomach(path)
```

For example, to restrict to specific years, countries, or ecoregions:

```{r download-subset}
download_stomach(path, year = 2000:2010, country = c("DK", "NO", "SE"))
download_stomach(path, ecoregion = "Greater North Sea")
```

In this vignette we run the pipeline on raw CSVs bundled with the package in `inst/extdata/`, as if you had downloaded them yourself. They were produced with, seen [here](https://github.com/maxlindmark/stomachr/blob/master/data-raw/build_example_data.R):

```{r download-example}
download_stomach(path, year = 2020:2024, ecoregion = "Greater North Sea")
```

```{r load-extdata, eval = TRUE}
library(stomachr)
path <- system.file("extdata", package = "stomachr")
```

### Step 2: `join_stomach_data()`

The next step is to join the four CSVs. This means attaching haul and file metadata to predators, classifying each stomach as `"food"`, `"empty"`, or `"unidentified"`, and optionally imputes missing coordinates from ICES rectangle midpoints (defaults to `"food"`).

<span style="color:red;">**Known country-specific unit issues** (currently requires manual correction before proceeding). All five fixes below are worked examples in the [known issues vignette](https://maxlindmark.github.io/stomachr/articles/known-issues.html) (`vignette("known-issues", package = "stomachr")` locally).

- Belgium (`BE`): `pred_length` is in mm -- divide by 10
- Denmark (`DK`): `ind_wgt` is in kg -- multiply by 1000
- Sweden, one upload (`tbl_upload_id == "8337"`, cod, Baltic Sea): `pred_length` is in mm -- divide by 10 (not a country-wide issue like the two above -- Sweden's other uploads are fine)
- Same Sweden upload: uses a legacy `regurgitated` code confirmed by ICES stomach analysts (`1` = intact, `2` = regurgitated, instead of the usual `0`/`NA` = not regurgitated, `1` = regurgitated) -- shift it by `-1` before `drop_invalid()`, or it wrongly discards good records
- Norway (`NO`): `number` is populated above `1` on many records despite each one being a single individually-measured fish (own `Length`, own `IndWgt`, `Regurgitated`/`StomachEmpty` capped at 0/1, prey-weight ratios that get *worse* when divided by `Number`) -- treated as a data-entry error, not real pooling; force it to `1`</span>

These things are not fixed internally in the join function, in case the database gets corrected.

This following error is though corrected internally, because ices rectangle is filled in. Some submissions (seen in Celtic Seas data) have corrupt `ICESrectangle` codes (e.g. `"46E9"` stored as `"46000000000"`, as if Excel read it as scientific notation). `join_stomach_data()` detects codes that don't match the real `##L#` format and recomputes them from `ShootLat`/`ShootLong`. How many rows were affected are printed by the function.

```{r unit-fix, eval = TRUE}
dat <- join_stomach_data(path)

dat <- dat |>
  dplyr::mutate(
    pred_length  = dplyr::if_else(country == "BE" | tbl_upload_id == "8337", pred_length / 10, pred_length),
    ind_wgt      = dplyr::if_else(country == "DK", ind_wgt * 1000, ind_wgt),
    regurgitated = dplyr::if_else(tbl_upload_id == "8337", regurgitated - 1, regurgitated),
    number       = dplyr::if_else(country == "NO", 1, number)
  )
```

### Step 3: `add_taxonomy()`

Joins scientific names and higher taxonomy (class, order, family, phylum) for both predators and prey from the internal WoRMS lookup. Prey with `aphia_id_prey = NA` in non-empty stomachs are labelled `"Unknown"` so their weight is not silently lost.

```{r taxonomy, eval = TRUE}
dat <- add_taxonomy(dat)
```

### Step 4: `unpool_predators()`

Some `PredatorInformation` records have `Number > 1`: not one fish, but a pooled group of that many ([file format documentation](https://datsu.ices.dk/web/selRep.aspx?Dataset=157): "Number of specimens taken for stomach analyses (pooled samples)". If you want to look at prey weigths in stomachs, e.g. for feeding rates, it's important that they are filtered or expanded, so that 1 row = 1 individual predator.

- `method = "uncount"` (default): expands each pooled record into `Number` rows (one pseudo-individual per implied fish), distributing `count`/`weight`/`regurgitated` so they sum back to the original totals (Example: for `Number = 10, Regurgitated = 3`, 3 of the 10 copies get `regurgitated = 1`, so `drop_invalid()` below drops exactly those 3, not all 10 or none). Note, these aren't independent observations, so treat them accordingly for anything computing per-individual variance.
- `method = "filter"`: drops `Number > 1` records outright instead.

This has to run before `drop_invalid()`, not after -- otherwise a partially-regurgitated pooled group gets dropped (or kept) as a whole before it can be resolved per implied individual.

```{r unpool, eval = TRUE}
dat <- unpool_predators(dat, method = "uncount")
```

### Step 5: `drop_invalid()`

Removes predators with `regurgitated >= 1`. Stomach contents of regurgitated fish are not really usable. The `na_regurgitated` argument controls whether `NA` values are treated as not regurgitated (`"keep"`, default) or regurgitated (`"drop"`).

```{r drop, eval = TRUE}
dat <- drop_invalid(dat, na_regurgitated = "keep")
```

### Step 6: `impute_size()`

Estimates missing prey weight and/or length via L/W parameters or observed means, and estimates missing predator weight from length. Also creates the final `predator_weight` column (observed weight if available, otherwise estimated from length). L/W parameters are looked up from the internal table (FishBase for fish, Robinson 2010 for invertebrates) with hierarchical fallback: species -> family -> order -> class -> phylum -> universal (a = 0.01, b = 3).

Future versions of this package could fit mixed models instead and impute weights in a more clever way, accounting also for size, etc!

- `which`: impute `"prey"`, `"pred"`, or `"both"` (default)
- `method`: `"lw_params"` (default) uses the bundled L/W table; `"observations"` uses mean sizes from other records of the same species
- `size`: impute `"weight"`, `"length"`, or `"both"` (default)
- `fill_if_no_size`: if `TRUE` (default), prey with no recorded size borrow from other records of the same species before applying L/W

```{r impute, eval = TRUE}
dat <- impute_size(dat, which = "both", method = "lw_params", size = "both")
```

### Step 7: `trim_data()`

Returns only the analysis-ready columns.

```{r trim, eval = TRUE}
dat <- trim_data(dat)
```

### Steps 8--9: `sense_check()` and `drop_flagged()`

`sense_check()` adds a `sense_flag` column marking implausible records. `drop_flagged()` removes them. You can inspect flagged records before dropping by examining `tbl_predator_information_id` in the raw data.

```{r check, eval = TRUE}
dat <- sense_check(dat)
dat <- drop_flagged(dat)
```
