---
title: "Example analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette picks up where `vignette("example-workflow", package = "stomachr")` leaves off, using the same bundled North Sea data (2020--2024). See that vignette for what each pipeline step does; here we just reconstruct the cleaned `dat` in one go and go straight to some example plots.

```{r load-extdata, eval = TRUE}
library(stomachr)
path <- system.file("extdata", package = "stomachr")
```

```{r rebuild-dat, eval = TRUE}
dat <- join_stomach_data(path) |>
  dplyr::mutate(
    pred_length  = dplyr::if_else(country == "BE" | tbl_upload_id == "8337", pred_length / 10, pred_length),
    ind_wgt      = dplyr::if_else(country == "DK", ind_wgt * 1000, ind_wgt),
    regurgitated = dplyr::if_else(tbl_upload_id == "8337", regurgitated - 1, regurgitated),
    number       = dplyr::if_else(country == "NO", 1, number)
  ) |>
  add_taxonomy() |>
  unpool_predators() |>
  drop_invalid() |>
  impute_size() |>
  trim_data() |>
  sense_check() |>
  drop_flagged()
```

The next sections show some basic plots for the most common predators in the build in North Sea data (2020--2024).

```{r setup-plots, eval = TRUE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)

theme_set(theme_minimal())
n <- 10

top <- dat |>
  distinct(tbl_predator_information_id, .keep_all = TRUE) |>
  count(predator_scientific_name, sort = TRUE) |>
  slice_head(n = n) |>
  pull(predator_scientific_name)
```

### Sampling locations

```{r plot-map, eval = TRUE, message = FALSE, warning = FALSE}
plot_map(dat, color = "year")
```

### Number of empty, food, and unidentified stomachs by year and predators

For this plot, we use `distinct()` to get one row per predator, the get the frequency of stomach-statuses across years and predator species.

```{r plot-status, eval = TRUE, fig.height = 8}
dat |>
  filter(predator_scientific_name %in% top) |>
  distinct(tbl_predator_information_id, .keep_all = TRUE) |>
  count(year, predator_scientific_name, stomach_status) |>
  ggplot(aes(year, n, fill = stomach_status)) +
  geom_col(alpha = 0.9) +
  scale_x_continuous(breaks = \(x) pretty(x, n = 3)) +
  facet_wrap(~predator_scientific_name, ncol = 3, scales = "free_y") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  labs(x = "Year", y = "Number of predators", fill = "Stomach status")
```

### Predator-prey mass ratio (PPMR)

To plot predator-prey mass ratios, we can use the final dataset, but we need to `uncount(count)`, such that one row is one individual prey!

```{r plot-ppmr, eval = TRUE, fig.height = 10}
dat |>
  filter(predator_scientific_name %in% top, stomach_status == "food") |>
  drop_na(predator_weight, prey_weight_ind) |>
  drop_na(prey_class) |>
  uncount(count) |>
  ggplot(aes(predator_weight, prey_weight_ind, color = prey_class)) +
  facet_wrap(~predator_scientific_name, scales = "free", ncol = 3) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  scale_x_log10() +
  scale_y_log10() +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Predator weight (g)", y = "Prey weight (g)", color = "Prey class") +
  theme(legend.position = "bottom")
```

### Diet composition by prey class

```{r plot-diet-class, eval = TRUE}
plot_dat <- dat |>
  filter(
    predator_scientific_name %in% top,
    stomach_status == "food",
    !is.na(prey_weight_all_ind)
  ) |>
  mutate(
    prey_group = fct_lump_n(coalesce(prey_class, "Other"), n = 5, w = prey_weight_all_ind, other_level = "Other"),
    n_pred = n_distinct(tbl_predator_information_id), .by = predator_scientific_name
  ) |>
  summarise(
    total_weight = sum(prey_weight_all_ind),
    n_pred = first(n_pred),
    .by = c(predator_scientific_name, prey_group)
  ) |>
  mutate(frac = total_weight / sum(total_weight), .by = predator_scientific_name)

ggplot(plot_dat, aes(frac, predator_scientific_name, fill = prey_group)) +
  geom_col() +
  geom_text(
    data = distinct(plot_dat, predator_scientific_name, n_pred),
    aes(x = 1.01, y = predator_scientific_name, label = n_pred),
    hjust = 0, inherit.aes = FALSE, size = 3
  ) +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_continuous(limits = c(0, 1.1), expand = c(0, 0)) +
  coord_cartesian(expand = FALSE, clip = "off") +
  labs(x = "Diet proportion (by weight)", y = NULL, fill = "Prey class")
```

### Cod diet over time by prey family

```{r plot-cod, eval = TRUE}
dat |>
  filter(
    predator_scientific_name == "Gadus morhua",
    stomach_status == "food",
    !is.na(prey_weight_all_ind)
  ) |>
  mutate(prey_group = fct_lump_n(prey_family, n = n, w = prey_weight_all_ind, other_level = "Other")) |>
  summarise(total_weight = sum(prey_weight_all_ind), .by = c(year, prey_group)) |>
  mutate(frac = total_weight / sum(total_weight), .by = year) |>
  ggplot(aes(year, frac, fill = prey_group)) +
  geom_col() +
  scale_fill_brewer(palette = "Spectral", na.value = "grey80") +
  scale_x_continuous(breaks = \(x) pretty(x, n = 6)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    x = "Year", y = "Diet proportion (by weight)", fill = "Prey family",
    title = "Gadus morhua"
  )
```

## References

Froese, R. and D. Pauly (eds.) FishBase. World Wide Web electronic publication. www.fishbase.org.

Robinson, R.A. et al. (2010). Trophic relationships of marine benthic invertebrates in the North Sea. *Journal of the Marine Biological Association of the United Kingdom*, 90(7), 1375-1388.
