---
title: "Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`stomachr` provides a tidy pipeline for downloading, reading, cleaning, and preparing ICES stomach content data for various diet analysis. The data come from the [ICES stomach content database](https://stomachdata.ices.dk) and follow the new four-table exchange format, joined by `tblUploadID`, `tblHaulID`, and `tblPredatorInformationID`.

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("maxlindmark/stomachr")
```

## Workflow

### Step 1: Download

You can download the data from the portal and go to Step 2. Or you can call `download_stomach()`, which downloads data from the ICES API and writes the four CSVs to a local directory (which you control via `path`). You can filter by `year`, `country`, and `ecoregion` (one of `"Baltic Sea"`, `"Celtic Seas"`, `"Greater North Sea"`).

```{r download}
path <- "data/raw"
download_stomach(path)
```

For example, to restrict to specific years, countries, or ecoregions:

```{r download-subset}
download_stomach(path, year = 2000:2010, country = c("DK", "NO", "SE"))
download_stomach(path, ecoregion = "Greater North Sea")
```

In this vignette we run the pipeline on raw CSVs bundled with the package in `inst/extdata/`, as if you had downloaded them yourself. They were produced with, seen [here](https://github.com/maxlindmark/stomachr/blob/master/data-raw/build_example_data.R):

```{r download-example}
download_stomach(path, year = 2020:2024, ecoregion = "Greater North Sea")
```

```{r load-extdata, eval = TRUE}
library(stomachr)
path <- system.file("extdata", package = "stomachr")
```

### Step 2: `join_stomach_data()`

The next step is to join the four CSVs. This means attaching haul and file metadata to predators, classifying each stomach as `"food"`, `"empty"`, or `"unidentified"`, and optionally imputes missing coordinates from ICES rectangle midpoints (defaults to `"food"`).

<span style="color:red;">**Known country-specific unit issues** (currently requires manual correction before proceeding). All five fixes below are worked examples in the [known issues vignette](https://maxlindmark.github.io/stomachr/articles/known-issues.html) (`vignette("known-issues", package = "stomachr")` locally).

- Belgium (`BE`): `pred_length` is in mm -- divide by 10
- Denmark (`DK`): `ind_wgt` is in kg -- multiply by 1000
- Sweden, one upload (`tbl_upload_id == "8337"`, cod, Baltic Sea): `pred_length` is in mm -- divide by 10 (not a country-wide issue like the two above -- Sweden's other uploads are fine)
- Same Sweden upload: uses a legacy `regurgitated` code confirmed by ICES stomach analysts (`1` = intact, `2` = regurgitated, instead of the usual `0`/`NA` = not regurgitated, `1` = regurgitated) -- shift it by `-1` before `drop_invalid()`, or it wrongly discards good records
- Norway (`NO`): `number` is populated above `1` on many records despite each one being a single individually-measured fish (own `Length`, own `IndWgt`, `Regurgitated`/`StomachEmpty` capped at 0/1, prey-weight ratios that get *worse* when divided by `Number`) -- treated as a data-entry error, not real pooling; force it to `1`</span>

These things are not fixed internally in the join function, in case the database gets corrected.

This following error is though corrected internally, because ices rectangle is filled in. Some submissions (seen in Celtic Seas data) have corrupt `ICESrectangle` codes (e.g. `"46E9"` stored as `"46000000000"`, as if Excel read it as scientific notation). `join_stomach_data()` detects codes that don't match the real `##L#` format and recomputes them from `ShootLat`/`ShootLong`. How many rows were affected are printed by the function.

```{r unit-fix, eval = TRUE}
dat <- join_stomach_data(path)

dat <- dat |>
  dplyr::mutate(
    pred_length  = dplyr::if_else(country == "BE" | tbl_upload_id == "8337", pred_length / 10, pred_length),
    ind_wgt      = dplyr::if_else(country == "DK", ind_wgt * 1000, ind_wgt),
    regurgitated = dplyr::if_else(tbl_upload_id == "8337", regurgitated - 1, regurgitated),
    number       = dplyr::if_else(country == "NO", 1, number)
  )
```

### Step 3: `add_taxonomy()`

Joins scientific names and higher taxonomy (class, order, family, phylum) for both predators and prey from the internal WoRMS lookup. Prey with `aphia_id_prey = NA` in non-empty stomachs are labelled `"Unknown"` so their weight is not silently lost.

```{r taxonomy, eval = TRUE}
dat <- add_taxonomy(dat)
```

### Step 4: `drop_invalid()`

Some `PredatorInformation` records have `Number > 1`: not one fish, but a pooled sample of that many ([file format documentation](https://datsu.ices.dk/web/selRep.aspx?Dataset=157): "Number of specimens taken for stomach analyses (pooled samples)"). `Regurgitated` and `StomachEmpty` are counts out of `Number`. `drop_invalid()` resolves regurgitation per record and adds two columns:

- `n_stomachs = number - regurgitated`: the usable stomachs behind the record (`number = NA` is treated as `1`). Records with `n_stomachs <= 0` are dropped. For a single fish that's the same as dropping `regurgitated >= 1`, but a pool of 10 with 3 regurgitated is kept with `n_stomachs = 7`, and its prey is attributed to those 7.
- `n_empty`: how many of the `n_stomachs` were empty. It equals `n_stomachs` when the record has no prey, is `0` for a single fish with prey, and otherwise comes from `StomachEmpty` (capped at `n_stomachs - 1`, since a record with prey had at least one fed fish). It is `NA` for a pooled record with prey but no `StomachEmpty` reported. `stomach_status` can't express "3 of 10 empty" on a pooled record, so use `n_empty / n_stomachs` for proportions of empty stomachs.

The `na_regurgitated` argument controls whether `NA` values are treated as not regurgitated (`"keep"`, default) or drop the record (`"drop"`).

```{r drop, eval = TRUE}
dat <- drop_invalid(dat, na_regurgitated = "keep")
```

#### Using `n_stomachs`

Pooled records stay one row each, holding the combined prey of `n_stomachs` fish. Anything per fish needs `n_stomachs` in the denominator:

- Mean stomach content: `sum(weight) / sum(n_stomachs)`, not `mean()` over records.
- Counts or prey weight in a model with a log link (Poisson, negative binomial, Tweedie, Gamma): add `offset(log(n_stomachs))`. For single fish the offset is 0. For Tweedie, the dispersion of a sum of n fish also scales, by roughly n^(1-p), so pooled records carry slightly different variance.
- Presence/absence of a prey: a pooled record answers "did any of n fish eat it", i.e. 1 - (1 - p)^n. Use a binomial model with a cloglog link and `offset(log(n_stomachs))`, not a logit link.
- Stomach fullness: the predator size on a pooled record describes one representative fish, so divide by `n_stomachs * predator_weight` (or use `offset(log(n_stomachs) + log(predator_weight))`).
- Diet composition (proportions of prey): no offset needed; a pool is a noisier sample of the same proportions.

### Step 5 (optional): `unpool_predators()`

If you need one row per fish instead, `unpool_predators()` expands each pooled record into `n_stomachs` synthetic individuals: `n_stomachs - n_empty` fed copies share the prey, and `n_empty` copies get no prey and `stomach_status = "empty"`. `count` is split as evenly as possible (`count = 7` over 3 copies becomes `3, 2, 2`) and `weight` follows, so totals sum back to the original.

That even split is an assumption, not data: the real split could have been `7, 0, 0`. Totals and means are unaffected, but the copies look far more alike than real fish do, so between-fish variance is understated, the number of independent observations is overstated, and frequency of occurrence is biased upward (`count = 3` over 3 fish becomes 100% occurrence, when the truth could be 33%). Prefer `n_stomachs` for modelling.

```{r unpool, eval = FALSE}
dat <- unpool_predators(dat)
```

### Step 6: `impute_size()`

Estimates missing prey weight and/or length via L/W parameters or observed means, and estimates missing predator weight from length. Also creates the final `predator_weight` column (observed weight if available, otherwise estimated from length). L/W parameters are looked up from the internal table (FishBase for fish, Robinson 2010 for invertebrates) with hierarchical fallback: species -> family -> order -> class -> phylum -> universal (a = 0.01, b = 3).

Future versions of this package could fit mixed models instead and impute weights in a more clever way, accounting also for size, etc!

- `which`: impute `"prey"`, `"pred"`, or `"both"` (default)
- `method`: `"lw_params"` (default) uses the bundled L/W table; `"observations"` uses mean sizes from other records of the same species
- `size`: impute `"weight"`, `"length"`, or `"both"` (default)
- `fill_if_no_size`: if `TRUE` (default), prey with no recorded size borrow from other records of the same species before applying L/W

```{r impute, eval = TRUE}
dat <- impute_size(dat, which = "both", method = "lw_params", size = "both")
```

### Step 7: `trim_data()`

Returns only the analysis-ready columns.

```{r trim, eval = TRUE}
dat <- trim_data(dat)
```

### Steps 8--9: `sense_check()` and `drop_flagged()`

`sense_check()` adds a `sense_flag` column marking implausible records. `drop_flagged()` removes them. You can inspect flagged records before dropping by examining `tbl_predator_information_id` in the raw data.

```{r check, eval = TRUE}
dat <- sense_check(dat)
dat <- drop_flagged(dat)
```
