---
title: "Known issues"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Known issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Here I track known issues with the database. These include data entry errors, which `stomachr` does not fix internally, but show how to fix in the [example-workflow](https://maxlindmark.github.io/stomachr/articles/example-workflow.html), and ambiguities in the [file format documentation](https://datsu.ices.dk/web/selRep.aspx?Dataset=157) (which if fixed, can reduce data entry errors).

First download data.

```{r download, eval = TRUE, message = FALSE}
library(stomachr)
library(dplyr)
library(ggplot2)

theme_set(theme_minimal())

regions <- c("Baltic Sea", "Celtic Seas", "Greater North Sea")

pred <- lapply(regions, function(r) {
  path <- tempfile()
  download_stomach(path, ecoregion = r)

  fi <- readr::read_csv(file.path(path, "File_information.csv"), show_col_types = FALSE) |>
    janitor::clean_names() |>
    transmute(tbl_upload_id = as.character(tbl_upload_id), country = as.character(country))

  prey_n <- readr::read_csv(file.path(path, "PreyInformation.csv"), show_col_types = FALSE) |>
    janitor::clean_names() |>
    transmute(tbl_predator_information_id = as.character(tbl_predator_information_id)) |>
    count(tbl_predator_information_id, name = "n_prey_rows")

  readr::read_csv(file.path(path, "PredatorInformation.csv"), show_col_types = FALSE) |>
    janitor::clean_names() |>
    transmute(
      tbl_upload_id = as.character(tbl_upload_id),
      tbl_predator_information_id = as.character(tbl_predator_information_id),
      tbl_haul_id = as.character(tbl_haul_id),
      fish_id = as.character(fish_id),
      aphia_id_predator = as.numeric(aphia_id_predator),
      number = as.numeric(number),
      measurement_increment = as.numeric(measurement_increment),
      length = as.numeric(length),
      ind_wgt = as.numeric(ind_wgt),
      regurgitated = as.numeric(regurgitated),
      year = as.numeric(year),
      ecoregion = r
    ) |>
    left_join(fi, by = "tbl_upload_id") |>
    left_join(prey_n, by = "tbl_predator_information_id") |>
    mutate(has_prey = !is.na(n_prey_rows) & n_prey_rows > 0)
}) |>
  bind_rows()
```

## Known data issues

Units of length and weight of predators. For prey, there are columns describing which unit it is. For predators, it's in the description for `IndWgt` and `Length`. `stomachr` flags extreme outliers in the `join_stomach_data()`, based on residuals of generic length-weight parameters.

```{r lw-setup, eval = TRUE}
lw_a <- 0.01
lw_b <- 3

lw_dat <- pred |>
  filter(number == 1, !is.na(length), length > 0, !is.na(ind_wgt), ind_wgt > 0) |>
  mutate(
    log_ratio = log10(ind_wgt / (lw_a * length^lw_b)),
    flagged = abs(log_ratio) > 1.5
  )

lw_curve <- data.frame(length = seq(1, 200, length.out = 200)) |>
  mutate(ind_wgt = lw_a * length^lw_b)
```

Here we see deviations from an isometric length-weight curve (`weight = a * length^b`, `a = 0.01, b = 3`), per record (usually a single predator, but can be a pooled predator as well if `number` > 1). A record now gets flagged if `abs(log_ratio) > 1.5`.

```{r flag-diagnostic, eval = TRUE, fig.width = 7, fig.height = 4}
lw_dat |>
  ggplot(aes(log_ratio, fill = country)) +
  geom_histogram(binwidth = 0.1) +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "#D95F02") +
  scale_fill_brewer(palette = "Dark2")
```

```{r flag-scatter, eval = TRUE, fig.width = 8, fig.height = 5, warning = FALSE}
lw_dat |>
  ggplot(aes(length, ind_wgt, color = flagged)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "#D95F02"))
```

```{r flag-table, eval = TRUE}
lw_dat |> filter(flagged) |> count(country, sort = TRUE)
```

We don't know yet whether a flagged record's `length` or `ind_wgt` is the wrong one. Length is easier to work with. Assuming all countries do the same error (doesn't have to be though), we can see if a length "correction" leads to implausible lengths (if length varies between 10 and 100, it cannot be in mm because we never collect 1 cm fish. The similar test is harder to make for weight).

```{r lw-corrected, eval = TRUE}
call_by_record <- lw_dat |>
  filter(flagged) |>
  mutate(call = if_else(length / 10 < 5, "weight", "length"))

call_by_country <- call_by_record |>
  count(country, call) |>
  slice_max(n, by = country) |>
  select(country, call)

lw_dat_corrected <- lw_dat |>
  left_join(call_by_country, by = "country") |>
  mutate(
    length = if_else(flagged & call == "length", length / 10, length),
    ind_wgt = if_else(flagged & call == "weight", ind_wgt * 1000, ind_wgt)
  )
```

Now look at the corrected records:

### 1. Belgium: predator length recorded in mm, not cm

```{r be-diagnostic, eval = TRUE, fig.width = 8, fig.height = 5, warning = FALSE}
be_ref <- lw_dat_corrected |> filter(country != "BE")
be_points <- bind_rows(
  lw_dat |> filter(country == "BE") |> mutate(group = "BE, as recorded"),
  lw_dat |> filter(country == "BE", flagged) |> mutate(length = length / 10, group = "BE, length / 10")
)

ggplot() +
  geom_point(data = be_ref, aes(length, ind_wgt), color = "grey70", alpha = 0.3, size = 0.8) +
  geom_line(data = lw_curve, aes(length, ind_wgt), linetype = "dashed", linewidth = 0.5) +
  geom_point(data = be_points, aes(length, ind_wgt, color = group), alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("BE, as recorded" = "#D95F02", "BE, length / 10" = "#1B9E77")) +
  labs(
    x = "Predator length, cm", y = "Predator weight, g", color = NULL,
    caption = "Belgium only lands on the curve once length is divided by 10",
    subtitle = "Grey = every other country's records, any species"
  )
```

- **Fix applied in the example vignettes**: `pred_length = if_else(country == "BE", pred_length / 10, pred_length)`.

### 2. Denmark: predator weight recorded in kg, not g

```{r dk-diagnostic, eval = TRUE, fig.width = 8, fig.height = 5, warning = FALSE}
dk_ref <- lw_dat_corrected |> filter(country != "DK")
dk_points <- bind_rows(
  lw_dat |> filter(country == "DK") |> mutate(group = "DK, as recorded"),
  lw_dat |> filter(country == "DK", flagged) |> mutate(ind_wgt = ind_wgt * 1000, group = "DK, weight x 1000")
)

ggplot() +
  geom_point(data = dk_ref, aes(length, ind_wgt), color = "grey70", alpha = 0.3, size = 0.8) +
  geom_line(data = lw_curve, aes(length, ind_wgt), linetype = "dashed", linewidth = 0.5) +
  geom_point(data = dk_points, aes(length, ind_wgt, color = group), alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("DK, as recorded" = "#D95F02", "DK, weight x 1000" = "#1B9E77")) +
  labs(
    x = "Predator length, cm", y = "Predator weight, g", color = NULL,
    caption = "Denmark only lands on the curve once weight is multiplied by 1000",
    subtitle = "Grey = every other country's records, any species"
  )
```

- **Fix applied in the example vignettes**: `ind_wgt = if_else(country == "DK", ind_wgt * 1000, ind_wgt)`.

### 3. Sweden: predator length recorded in mm, not cm (*one upload only)

```{r se-diagnostic, eval = TRUE, fig.width = 8, fig.height = 5, warning = FALSE}
se_ref <- lw_dat_corrected |> filter(!(country == "SE" & flagged))
se_points <- bind_rows(
  lw_dat |> filter(country == "SE", flagged) |> mutate(group = "SE, as recorded"),
  lw_dat |> filter(country == "SE", flagged) |> mutate(length = length / 10, group = "SE, length / 10")
)

ggplot() +
  geom_point(data = se_ref, aes(length, ind_wgt), color = "grey70", alpha = 0.3, size = 0.8) +
  geom_line(data = lw_curve, aes(length, ind_wgt), linetype = "dashed", linewidth = 0.5) +
  geom_point(data = se_points, aes(length, ind_wgt, color = group), alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("SE, as recorded" = "#D95F02", "flagged SE, length / 10" = "#1B9E77")) +
  labs(
    x = "Predator length, cm (log10)", y = "Predator weight, g (log10)", color = NULL,
    caption = "Sweden's flagged records only land on the curve once length is divided by 10",
    subtitle = "Grey = every other unflagged record, any country/species"
  )
```

- Only `country == "SE" & flagged` is affected. Note! In contrast to BE and DK, this is limited to a single upload (`tbl_upload_id == "8337"`, cod, Baltic Sea), not for all submissions by the country.
- **Fix applied in the example vignettes**: `pred_length = if_else(country == "BE" | tbl_upload_id == "8337", pred_length / 10, pred_length)`, same line as the Belgium fix above, since both are the same length-in-mm issue.

### 4. Sweden: `Regurgitated` sometimes uses a legacy code, not a count

This is a tricky one. 

After consultation with ICES stomach analysts directly, the normal convention seem to be that `0`/`NA` = not regurgitated, `1` = regurgitated. This does not match the [instructions](https://datsu.ices.dk/web/selRep.aspx?Dataset=157), which says: "Number of stomachs regurgitated", i.e., a count. Nor does it match the old version in Sweden's national database, where `1` = intact, `2` = regurgitated. 

Hence, a `1` in Sweden could therefore mean that it was regurgitated (if the new convention is used), or that it was not regurgitated (old Swedish convention). `2` in a Swedish record could mean 2 regurgitated stomachs (i.e., against convention, but in line with the actual instructions), **but only if** the record is a pooled record. *However*, Sweden does not normally report pooled stomachs. We can see that's not the case here, because even if records with `2` from Sweden did refer to the correct input of number of regurgitated stomachs, then this number should never exceed the value in `number`. A better test is this: if the old code is used, then there should be no zeroes in the `tbl_upload_id`, which if the new code is used (under whichever interpretation of it), should be very unlikely!

(There are other issues with regurgitation, see below.)

```{r regurg-math-check, eval = TRUE}
# Regurgitated is a count out of Number (ICES: "Number of stomachs regurgitated")
# So Regurgitated > Number is not impossible unless a different code is in use.
pred |>
  filter(!is.na(regurgitated), !is.na(number), regurgitated > number) |>
  count(tbl_upload_id, country)

pred |>
  filter(country == "SE") |>
  summarise(n = n(), .by = c(tbl_upload_id, regurgitated)) |>
  summarise(has_zero = any(regurgitated == 0, na.rm = TRUE), .by = tbl_upload_id) |>
  filter(!has_zero)
```

- **Fix applied in the example vignettes**: `regurgitated = dplyr::if_else(tbl_upload_id == "8337", regurgitated - 1, regurgitated)`

### 5. Corrupted `ICESrectangle` codes (handled automatically, no manual fix needed)

Some submissions (seen in Celtic Seas data) have corrupt `ICESrectangle` codes -- e.g. `"46E9"` stored as `"46000000000"`, as if a spreadsheet read it as scientific notation before export. Unlike everything else in this section, `join_stomach_data()` already handles this on its own: it detects codes that don't match the real `##L#` format and recomputes them from `ShootLat`/`ShootLong`, printing how many rows were affected.

```{r ices-rectangle-check, eval = TRUE, message = FALSE}
valid_rect <- "^[0-9]{2}[A-Za-z][0-9]$"

rect_dat <- lapply(regions, function(r) {
  path <- tempfile()
  download_stomach(path, ecoregion = r)
  readr::read_csv(file.path(path, "HaulInformation.csv"),
    col_types = readr::cols(ICESrectangle = readr::col_character()), show_col_types = FALSE
  ) |>
    janitor::clean_names() |>
    transmute(ices_rectangle = ice_srectangle, ecoregion = r)
}) |>
  bind_rows()

rect_dat |>
  filter(!is.na(ices_rectangle), !grepl(valid_rect, ices_rectangle)) |>
  count(ices_rectangle, sort = TRUE)
```

- Malformed codes are consistently long, all-numeric strings -- the scientific-notation signature -- not random corruption.
- **No fix needed here beyond what's already in** `join_stomach_data()`; included for completeness, since every other entry in this section needs a manual workaround and this one doesn't.

### 6. Norway: `Number` is populated, but the record isn't actually pooled

`Number` can be greater than `1`, and the [file format documentation](https://datsu.ices.dk/web/selRep.aspx?Dataset=157) describes it as "Number of specimens taken for stomach analyses (pooled samples)" -- i.e. that record's prey rows *should* describe `Number` fish pooled together for all `n = number` predators that row corresponds to. It's unclear what the column `IndWgt` corresponds to when `Number` > 1

#### Which countries pool?

```{r pooling-by-country-year, eval = TRUE, message = FALSE, fig.width = 8, fig.height = 5, warning = FALSE}
pool_dat <- lapply(regions, function(r) {
  path <- tempfile()
  download_stomach(path, ecoregion = r)
  fi <- readr::read_csv(file.path(path, "File_information.csv"), show_col_types = FALSE) |>
    janitor::clean_names() |>
    transmute(tbl_upload_id = as.character(tbl_upload_id), country = as.character(country))
  readr::read_csv(file.path(path, "PredatorInformation.csv"), show_col_types = FALSE) |>
    janitor::clean_names() |>
    transmute(tbl_upload_id = as.character(tbl_upload_id), number = as.numeric(number), year = as.numeric(year)) |>
    left_join(fi, by = "tbl_upload_id")
}) |>
  bind_rows() |>
  filter(!is.na(number), !is.na(year))

pool_dat |>
  summarise(n = n(), pct_pooled = 100 * mean(number > 1), .by = c(country, year)) |>
  mutate(pooled = ifelse(pct_pooled == 0, "Not pooled", "pooled")) |>
  mutate(sum = sum(pct_pooled), .by = c(country)) |> 
  filter(sum > 0) |> 
  ggplot(aes(year, pct_pooled, size = n, color = country)) +
  geom_point(alpha = 0.8) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = NULL, y = "% of that country's records pooled that year", size = "N records", color = NULL)
```

- Only Norway and the Netherlands ever report `Number > 1`
- The Netherlands phased it out: ~74% of its 1980s-90s records were pooled, 0% in its 2020s submissions.
- Norway's only data here is 2020s, still ~66% has `Number > 1`.

#### Does dividing prey weight by `Number` help, or hurt?

A pooled record means that 1 row is not 1 individuals. Getting this right is crucial for calculating relative prey weight/stomach fullness. 

Here I illustrate this by comparing `total prey weight / predator weight` for records where we know 1 row = 1 individual (`Number == 1`), against pooled records left alone ("uncorrected") by row, and pooled records with prey weight divided by `Number` ("corrected"), one point per predator. 

Another insight here: Norway has `IndWgt` on its pooled records; the Netherlands doesn't (0 of 2,475 rows), so the Netherlands version uses the same universal length-weight curve from earlier (`a = 0.01, b = 3`) as the size reference instead. This is also a clue that Norway's number column may be mis-reported...

Every group below (individual, uncorrected, corrected) is restricted to the same set of species per country: those with *both* `Number == 1` and `Number > 1` records. Otherwise a species that's always pooled (or always individual) for that country would pull a whole group's ratios up or down for reasons that have nothing to do with pooling -- e.g. if the pooled species just happen to eat less on average, that would look like "pooling deflates the ratio" even with no pooling artefact at all.

```{r ratio-by-country-setup, eval = TRUE, message = FALSE}
prey_dat <- lapply(regions, function(r) {
  path <- tempfile()
  download_stomach(path, ecoregion = r)
  join_stomach_data(path) |>
    summarise(
      total_prey_weight = sum(weight, na.rm = TRUE),
      any_weight = any(!is.na(weight)),
      number = dplyr::first(number),
      ind_wgt = dplyr::first(ind_wgt),
      length = dplyr::first(pred_length),
      country = dplyr::first(country),
      aphia_id_predator = dplyr::first(aphia_id_predator),
      .by = tbl_predator_information_id
    )
}) |>
  bind_rows() |>
  filter(!is.na(number), any_weight, total_prey_weight > 0)
```

```{r ratio-data-prep, eval = TRUE, message = FALSE}
lw_a <- 0.01
lw_b <- 3

no_dat <- prey_dat |> filter(country == "NO", !is.na(ind_wgt), ind_wgt > 0)
no_pooled_species <- no_dat |> filter(number > 1) |> distinct(aphia_id_predator) |> pull()
no_indiv_species <- no_dat |> filter(number == 1) |> distinct(aphia_id_predator) |> pull()
no_species <- intersect(no_pooled_species, no_indiv_species)

nl_dat <- prey_dat |> filter(country == "NL", !is.na(length), length > 0) |>
  mutate(pred_wgt_from_length = lw_a * length^lw_b)
nl_pooled_species <- nl_dat |> filter(number > 1) |> distinct(aphia_id_predator) |> pull()
nl_indiv_species <- nl_dat |> filter(number == 1) |> distinct(aphia_id_predator) |> pull()
nl_species <- intersect(nl_pooled_species, nl_indiv_species)

# Restrict every group to aphia_id_predator %in% no_species/nl_species -- species
# with BOTH individual and pooled records for that country. Without this, a
# species that's e.g. always pooled (never recorded as number == 1) would
# contribute to the "pooled" boxes with no individual counterpart to compare
# against, and the country-level comparison would really be comparing two
# different sets of species, not the same species measured two ways.
ratio_dat_raw <- bind_rows(
  no_dat |> filter(number == 1, aphia_id_predator %in% no_species) |> mutate(country = "Norway", group = "individual", ratio = total_prey_weight / ind_wgt),
  no_dat |> filter(number > 1, aphia_id_predator %in% no_species) |> mutate(country = "Norway", group = "pooled, uncorrected", ratio = total_prey_weight / ind_wgt),
  no_dat |> filter(number > 1, aphia_id_predator %in% no_species) |> mutate(country = "Norway", group = "pooled, corrected (/ Number)", ratio = (total_prey_weight / number) / ind_wgt),
  nl_dat |> filter(number == 1, aphia_id_predator %in% nl_species) |> mutate(country = "Netherlands", group = "individual", ratio = total_prey_weight / pred_wgt_from_length),
  nl_dat |> filter(number > 1, aphia_id_predator %in% nl_species) |> mutate(country = "Netherlands", group = "pooled, uncorrected", ratio = total_prey_weight / pred_wgt_from_length),
  nl_dat |> filter(number > 1, aphia_id_predator %in% nl_species) |> mutate(country = "Netherlands", group = "pooled, corrected (/ Number)", ratio = (total_prey_weight / number) / pred_wgt_from_length)
)
```

```{r ratio-plot, eval = TRUE, fig.width = 6, fig.height = 5, warning = FALSE}
ratio_raw <- ratio_dat_raw |>
  mutate(
    group_clean = dplyr::case_match(
      group,
      "individual" ~ "individual",
      "pooled, uncorrected" ~ "uncorrected",
      "pooled, corrected (/ Number)" ~ "corrected"
    ),
    group_clean = factor(group_clean, levels = c("individual", "uncorrected", "corrected"))
  )

ggplot(ratio_raw, aes(x = group_clean, y = ratio*100, color = group_clean)) +
  geom_jitter(position = position_jitter(width = 0.15, seed = 42), alpha = 0.4, size = 1) +
  geom_boxplot(aes(fill = group_clean), width = 0.4, alpha = 0.3, outlier.shape = NA, color = "black") +
  stat_summary(
    fun = median, geom = "text", color = "black", size = 3, vjust = -30,
    aes(label = signif(after_stat(y), 2))
  ) +
  scale_color_manual(values = c(individual = "grey30", corrected = "tomato", uncorrected = "steelblue2")) +
  scale_fill_manual(values = c(individual = "grey30", corrected = "tomato", uncorrected = "steelblue2")) +
  scale_y_continuous(trans = "sqrt") +
  facet_wrap(~country, scales = "free_y") +
  labs(
    x = NULL, y = "Prey weight as % of predator weight",
    title = "Does dividing pooled prey weight by Number land on the individual baseline?"
  ) +
  theme(legend.position = "none")
```

For the Netherlands (left): "corrected" lands is to "individual", while "uncorrected" is off by several-fold. Here number really is the number of fish pooled together. For Norway (right), "uncorrected" are more similar to "individual" than "corrected" does, but the uncorrected are for some reason slightly larger than the individual still. Dividing by `Number` however makes the difference between "individual" bigger.

#### `Regurgitated`/`StomachEmpty`: capped at 1 for Norway, scales up for the Netherlands

Lastly, to really check Norway here, if a row really is one fish, `Regurgitated` should never exceed 1 one fish can't have more than one regurgitated stomach.

```{r regurg-cap-check, eval = TRUE}
bind_rows(
  pred |> filter(country == "NO", number > 1) |> mutate(check = "Norway, Number > 1"),
  pred |> filter(country == "NL", number > 1) |> mutate(check = "Netherlands, Number > 1")
) |>
  summarise(
    n = n(),
    max_regurgitated = max(regurgitated, na.rm = TRUE),
    n_regurgitated_gt_1 = sum(regurgitated > 1, na.rm = TRUE),
    .by = check
  )
```

Norway: `Regurgitated` never exceeds 1, this is consistent with one fish per row. The Netherlands: it does exceed 1. 

```{r}
pred |>
  filter(country %in% c("NL", "NO"), number > 1, !is.na(regurgitated)) |>
  ggplot(aes(number, regurgitated)) +
  facet_wrap(~country) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_jitter(width = 0.3, height = 0.3, alpha = 0.4, size = 1) +
  labs(
    x = "Number", y = "Regurgitated",
    caption = "Regurgitated scales with Number in NL, not NO, as a real pooled count should",
    subtitle = "Dashed line = Regurgitated == Number (the upper bound for a genuine count)"
  )
```

Finally, let's check predator weights. If pooled, unclear what the `IndWgt` refers to.

```{r pooled-weight-diagnostic, eval = TRUE, fig.width = 9, fig.height = 7, warning = FALSE}
# lw_dat_corrected, not lw_dat: same reasoning as be_ref/dk_ref above -- if
# BE or DK happen to share a species with Norway's pooled records, their
# uncorrected values would otherwise contaminate this reference cloud too.
indiv_ref <- lw_dat_corrected |>
  add_count(aphia_id_predator, name = "n_indiv") |>
  filter(n_indiv >= 30)

pooled_species <- pred |>
  filter(number > 1, !is.na(ind_wgt), !is.na(length), ind_wgt > 0, length > 0) |>
  distinct(aphia_id_predator) |>
  inner_join(indiv_ref |> distinct(aphia_id_predator), by = "aphia_id_predator") |>
  pull(aphia_id_predator)

pooled <- pred |>
  filter(
    number > 1, !is.na(ind_wgt), !is.na(length), ind_wgt > 0, length > 0,
    aphia_id_predator %in% pooled_species
  )

cat("Countries contributing pooled records with a populated IndWgt:", 
    paste(unique(pooled$country), collapse = ", "), "\n")

pooled_points <- bind_rows(
  pooled |> mutate(plot_wgt = ind_wgt, assumption = "IndWgt, as recorded"),
  pooled |> mutate(plot_wgt = ind_wgt / number, assumption = "IndWgt / Number")
)

ggplot() +
  geom_point(
    data = indiv_ref |> filter(aphia_id_predator %in% pooled_species),
    aes(length, ind_wgt), color = "grey70", alpha = 0.3, size = 0.8
  ) +
  geom_point(data = pooled_points, 
             aes(length, plot_wgt, color = assumption), alpha = 0.8, size = 1.6) +
  facet_wrap(~aphia_id_predator, scales = "free") +
  scale_color_manual(values = c(
    "IndWgt, as recorded" = "#D95F02", "IndWgt / Number" = "#1B9E77")) +
  labs(
    x = "Length, cm", y = "Weight, g", color = NULL,
    caption = "Pooled records: does IndWgt or IndWgt / Number land on the individual cloud?",
    subtitle = "Grey = genuine individual (Number == 1) records, same species"
  ) +
  theme(legend.position = "bottom")
```

**What this means**: 
  - **Netherlands**: genuine pooling, matching the ICES description. `IndWgt` is never populated (nothing to report for a pooled group), `Regurgitated`/`StomachEmpty` scale with `Number` as real counts, and dividing prey weight by `Number` recovers a per-fish rate that matches genuine individuals.
  - **Norway**: not pooling. Each row is one individually-measured fish, indicated by `IndWgt` being populated and matching up with a single individual, `Regurgitated`/`StomachEmpty` are capped at 0/1 like any single-fish record, and dividing prey weight by `Number` makes the per-fish rate *worse*, not better. This really suggests its a data entry error. 
- **Fix applied in the example vignettes**: `number = dplyr::if_else(country == "NO", 1, number)`. This keeps Norway's already-one-fish-per-row records untouched by `unpool_predators()`, which after this fix only ever expands genuine Netherlands pooling.

## Limitations in the ICES format documentation

Field descriptions are from [datsu.ices.dk/web/selRep.aspx?Dataset=157](https://datsu.ices.dk/web/selRep.aspx?Dataset=157) (Stomach Content dataset, Predator Information record), current as of 2026-08-20.

### 1. `Number`: says "species", almost certainly means "specimens"

- *"Number of **species** taken for stomach analyses (pooled samples)"*.
- A record has exactly one unique `AphiaIDPredator`, so there's nothing for "number of species" to count within one record.
- **Correction**: "number of **specimens**"
- See also the error above

```{r number-diagnostic, eval = TRUE}
# every record has exactly one AphiaIDPredator, regardless of Number,
# confirming "species" can't be the intended plural here
pred |>
  filter(!is.na(aphia_id_predator)) |>
  summarise(
    n_records = n(),
    max_species_per_record = 1L, # AphiaIDPredator is a scalar column by construction
    n_pooled_records = sum(number > 1, na.rm = TRUE),
    pct_pooled = round(100 * n_pooled_records / n_records, 1)
  )
```

### 2. `FishID`: described as "unique", but isn't

- *"**Unique** fish identification number for predator."*
- **Wrong as written**: up to 4 different predator records can share an identical `FishID` within the same unique haul.

```{r fishid-diagnostic, eval = TRUE, fig.width = 7, fig.height = 4}
fishid_reuse <- pred |>
  filter(!is.na(fish_id)) |>
  count(tbl_haul_id, fish_id, name = "n_records_sharing_id")

cat(
  "n predator records:", nrow(pred), "\n",
  "n distinct FishID (global):", n_distinct(pred$fish_id), "\n",
  "(haul, FishID) combinations shared by >1 record:", sum(fishid_reuse$n_records_sharing_id > 1), "\n"
)

fishid_reuse |>
  count(n_records_sharing_id) |>
  ggplot(aes(x = factor(n_records_sharing_id), y = n)) +
  geom_col(fill = "#1B9E77") +
  scale_y_log10() +
  labs(
    x = "Records sharing the same (haul, FishID)", y = "Count (log10 scale)",
    caption = "FishID is not unique within a haul"
  )
```

```{r fishid-preddid-check, eval = TRUE}
# Confirm the duplicated FishID groups are still made of genuinely separate
# records: tblPredatorInformationID (the real row key) must stay distinct.
dup_groups <- fishid_reuse |> filter(n_records_sharing_id > 1)

pred |>
  inner_join(dup_groups, by = c("tbl_haul_id", "fish_id")) |>
  summarise(
    n_records = n(),
    n_distinct_pred_id = n_distinct(tbl_predator_information_id),
    .by = c(tbl_haul_id, fish_id)
  ) |>
  count(n_records, n_distinct_pred_id)
```

- **Not duplicate rows**: `n_records == n_distinct_pred_id` every time. `tblPredatorInformationID` (the real row key) stays distinct. It's specifically `FishID` that fails to be unique.

```{r fishid-country-diagnostic, eval = TRUE, fig.width = 7, fig.height = 4}
dup_hauls_by_country <- pred |>
  inner_join(dup_groups, by = c("tbl_haul_id", "fish_id")) |>
  distinct(tbl_haul_id, country) |>
  count(country, name = "n_dup_hauls")

haul_totals <- pred |> distinct(tbl_haul_id, country) |> count(country, name = "n_hauls")

dup_hauls_by_country |>
  left_join(haul_totals, by = "country") |>
  mutate(pct_hauls_affected = 100 * n_dup_hauls / n_hauls) |>
  ggplot(aes(x = reorder(country, pct_hauls_affected), y = pct_hauls_affected)) +
  geom_col(fill = "#D95F02") +
  coord_flip() +
  labs(
    x = NULL, y = "% of that country's hauls with a repeated FishID",
    title = "FishID reuse is concentrated in one country"
  )
```

- **Not spread evenly**: Sweden = 622 of 632 duplicated groups (98%); every other affected country (NL, DK, LV) is well under 1% of its own hauls.
- **96% of Sweden's own hauls** have a repeated `FishID`. Suggests a reporting pattern.
- **Correction**: Note sure. This column is not critical (we rely on `tblPredatorInformationID`), but should be corrected and clarified what the column means.

### 3. `Length` and `IndWgt`: undocumented for pooled records (when `Number > 1`)

- For `Number == 1`, both fields are unambiguous: the length/weight of that one fish. But the unit of length should be in the header, now it says "Length of specimen". The head for weight has this: "Weight of predator in grams". 
- For a pooled record, ICES says nothing about what either value represents:
  - `Length` (*"Length of specimen ... has to be specified in cm"*). How is this number calculated when there are multiple fish?
  - `IndWgt` (*"Weight of predator in grams"*) is this total across all `Number` fish, or one representative fish? Well, only Norway fills this with `Number` > 1, and we've shown the mis-reported the `Number` field. If it its clarified that this is individual predator weight/length, maybe this wouldn't happen. 

### 4. `Regurgitated` and `StomachEmpty`: inconsistent wording, and it's easy to mistake for a binary flag (and may be in practice to treat as binary?!)

- `StomachEmpty`: *"Number of empty stomachs **in the sample**"*.
- `Regurgitated`: *"Number of stomachs regurgitated"* (no "in the sample"). This is mainly a wording inconsistency.
- Both are counts relative to `Number`, consistent with `Number`'s own "(pooled samples)" note. It's worth stating explicitly, since the convention among many analysts seems to be that `Regurgitated` is a 0/1 flag rather than a count. Could it be that countries that **do not** upload pooled stomachs treat it as a binary? It doesn't really matter if all they do is upload non-pooled stomachs. But it helps understand the columns, and make the `unpool_predators()` function in this package easier to understand (where we create pseudo-individuals and remove as many as are regurgitated).

### 5. General points
- Should impossible combinations be flagged during upload? For example, check if `Regurgitated` > `Number` (would catch the Swedish legacy code), check if `IndWgt` > prey weight, should it be possible to populate `IndWgt` if `Number` > 1?

## Open question: should a regurgitated stomach have prey content at all?

It's also worth asking why a truly regurgitated stomach would ever reach species-level identification in the first place. Speaking to onboard samplers, if a stomach is visibly regurgitated, a new predator is selected. Many fish in this data are classified as regurgitated, but still have prey contents... Typically the issue with regurgitated stomachs is that we cannot separate them from truly empty stomachs. 

