---
title: "Getting started with stomachr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with stomachr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`stomachr` provides a tidy pipeline for downloading, reading, cleaning, and
preparing ICES stomach content data for various diet analysis. The
data come from the [ICES stomach content database](https://stomachdata.ices.dk)
and follow the new four-table exchange format, joined by `tblUploadID`,
`tblHaulID`, and `tblPredatorInformationID`.

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("maxlindmark/stomachr")
```

## Workflow

### Step 1: Download

You can download the data from the portal and go to Step 2. Or you can call
`download_stomach()`, which downloads data from the ICES API and writes the four CSVs
to a local directory (which you control via `path`). Ecoregion filtering is not
supported by the API (nor in the actual data it seems!), but you can filter geographically after 
joining using `lat`/`lon` or `ices_rectangle`.

```{r download}
path <- "data/raw"
download_stomach(path)
```

To restrict to specific years or countries:

```{r download-subset}
download_stomach(path, year = 2000:2010, country = c("DK", "NO", "SE"))
```

In this vignette we run the pipeline on the raw CSVs for the North Sea (2020-2024) bundled with the package
in `inst/extdata/`, as if you had downloaded them yourself.

```{r load-extdata, eval = TRUE}
library(stomachr)
path <- system.file("extdata", package = "stomachr")
```

### Step 2: `join_stomach_data()`

The next step is to join the four CSVs. This means attaching haul and file metadata 
to predators, classifying each stomach as `"food"`, `"empty"`, or `"unidentified"`, 
deduplicates exact-duplicate prey rows (present in some national submissions),
converts prey length from mm to cm, handles the count sentinel value (9999 = unknown
multiplicity), and optionally imputes missing coordinates from ICES rectangle
midpoints (defaults to yes).

<span style="color:red;">**Known country-specific unit issues** (currently requires manual correction
before proceeding):

- Belgium (`BE`): `pred_length` is in mm -- divide by 10
- Denmark (`DK`): `ind_wgt` is in kg -- multiply by 1000</span>

```{r unit-fix, eval = TRUE}
dat <- join_stomach_data(path)

dat <- dat |>
  dplyr::mutate(
    pred_length = dplyr::if_else(country == "BE", pred_length / 10, pred_length),
    ind_wgt     = dplyr::if_else(country == "DK", ind_wgt * 1000, ind_wgt)
  )
```

### Step 3: `drop_invalid()`

Removes predators with `regurgitated >= 1`. Stomach contents of regurgitated
fish are techically not usable. The `na_regurgitated` argument controls
whether `NA` values are treated as not regurgitated (`"keep"`, default) or
regurgitated (`"drop"`).

```{r drop, eval = TRUE}
dat <- drop_invalid(dat, na_regurgitated = "keep")
```

### Step 4: `add_taxonomy()`

Joins scientific names and higher taxonomy (class, order, family, phylum) for
both predators and prey from the internal WoRMS lookup. Prey with
`aphia_id_prey = NA` in non-empty stomachs are labelled `"Unknown"` so their
weight is not silently lost.

```{r taxonomy, eval = TRUE}
dat <- add_taxonomy(dat)
```

### Step 5: `impute_size()`

Estimates missing prey weight and/or length via L/W parameters or observed
means, and estimates missing predator weight from length. Also creates the
final `predator_weight` column (observed weight if available, otherwise
estimated from length). L/W parameters are looked up from the internal table
(FishBase for fish, Robinson 2010 for invertebrates) with hierarchical
fallback: species -> family -> order -> class -> phylum -> universal (a = 0.01, b = 3).

Future versions of this package could fit mixed models instead and impute weights in a more clever way, accounting also for size, etc!

- `which`: impute `"prey"`, `"pred"`, or `"both"` (default)
- `method`: `"lw_params"` (default) uses the bundled L/W table; `"observations"` uses mean sizes from other records of the same species
- `size`: impute `"weight"`, `"length"`, or `"both"` (default)
- `fill_if_no_size`: if `TRUE` (default), prey with no recorded size borrow from other records of the same species before applying L/W

```{r impute, eval = TRUE}
dat <- impute_size(dat, which = "both", method = "lw_params", size = "both")
```

### Step 6: `trim_data()`

Returns only the analysis-ready columns.

```{r trim, eval = TRUE}
dat <- trim_data(dat)
```

### Steps 7--8: `sense_check()` and `drop_flagged()`

`sense_check()` adds a `sense_flag` column marking implausible records.
`drop_flagged()` removes them. You can inspect flagged records before dropping
by examining `tbl_predator_information_id` in the raw data.

```{r check, eval = TRUE}
dat <- sense_check(dat)
dat <- drop_flagged(dat)
```

## Example analyses

The next sections show some basic plots for the most common predators in the build in
North Sea data (2020--2024).

```{r setup-plots, eval = TRUE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)

theme_set(theme_light())
n <- 10

top <- dat |>
  distinct(tbl_predator_information_id, .keep_all = TRUE) |>
  count(predator_scientific_name, sort = TRUE) |>
  slice_head(n = n) |>
  pull(predator_scientific_name)
```

### Sampling locations

```{r plot-map, eval = TRUE, message = FALSE, warning = FALSE}
plot_map(dat, color = "year")
```

### Proportion of empty, food, and unidentified stomachs by year and predators

For this plot, we use `distinct()` to get one row per predator, the count the frequency of stomach-statuses across years and predator species:

```{r plot-status, eval = TRUE, fig.height = 8}
dat |>
  filter(predator_scientific_name %in% top) |>
  distinct(tbl_predator_information_id, .keep_all = TRUE) |>
  count(year, predator_scientific_name, stomach_status) |>
  mutate(frac = n / sum(n), .by = c(year, predator_scientific_name)) |>
  ggplot(aes(year, frac, fill = stomach_status)) +
  geom_col(alpha = 0.9) +
  scale_x_continuous(breaks = \(x) pretty(x, n = 3)) +
  facet_wrap(~predator_scientific_name, ncol = 3) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  labs(x = "Year", y = "Percent", fill = "Stomach status")
```

### Predator-prey mass ratio (PPMR)

To plot predator-prey mass ratios, we can use the final dataset, but we need to 
`uncount(count)`, such that one row is one individual prey!

```{r plot-ppmr, eval = TRUE, fig.height = 10}
dat |>
  filter(predator_scientific_name %in% top, stomach_status == "food") |>
  drop_na(predator_weight, prey_weight_ind) |>
  drop_na(prey_class) |> 
  uncount(count) |>
  ggplot(aes(predator_weight, prey_weight_ind, color = prey_class)) +
  facet_wrap(~predator_scientific_name, scales = "free", ncol = 3) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  scale_x_log10() +
  scale_y_log10() +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Predator weight (g)", y = "Prey weight (g)", color = "Prey class") +
  theme(legend.position = "bottom")
```

### Diet composition by prey class

```{r plot-diet-class, eval = TRUE}
plot_dat <- dat |>
  filter(
    predator_scientific_name %in% top,
    stomach_status == "food",
    !is.na(prey_weight_all_ind)
  ) |>
  mutate(
    prey_group = fct_lump_n(coalesce(prey_class, "Other"), n = 5, w = prey_weight_all_ind, other_level = "Other"),
    n_pred = n_distinct(tbl_predator_information_id), .by = predator_scientific_name
  ) |>
  summarise(
    total_weight = sum(prey_weight_all_ind),
    n_pred = first(n_pred),
    .by = c(predator_scientific_name, prey_group)
  ) |>
  mutate(frac = total_weight / sum(total_weight), .by = predator_scientific_name)

ggplot(plot_dat, aes(frac, predator_scientific_name, fill = prey_group)) +
  geom_col() +
  geom_text(
    data = distinct(plot_dat, predator_scientific_name, n_pred),
    aes(x = 1.01, y = predator_scientific_name, label = n_pred),
    hjust = 0, inherit.aes = FALSE, size = 3
  ) +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_continuous(limits = c(0, 1.1), expand = c(0, 0)) +
  coord_cartesian(expand = FALSE, clip = "off") +
  labs(x = "Diet proportion (by weight)", y = NULL, fill = "Prey class")
```

### Cod diet over time by prey family

```{r plot-cod, eval = TRUE}
dat |>
  filter(
    predator_scientific_name == "Gadus morhua",
    stomach_status == "food",
    !is.na(prey_weight_all_ind)
  ) |>
  mutate(prey_group = fct_lump_n(prey_family, n = n, w = prey_weight_all_ind, other_level = "Other")) |>
  summarise(total_weight = sum(prey_weight_all_ind), .by = c(year, prey_group)) |>
  mutate(frac = total_weight / sum(total_weight), .by = year) |>
  ggplot(aes(year, frac, fill = prey_group)) +
  geom_col() +
  scale_fill_brewer(palette = "Spectral", na.value = "grey80") +
  scale_x_continuous(breaks = \(x) pretty(x, n = 6)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    x = "Year", y = "Diet proportion (by weight)", fill = "Prey family",
    title = "Gadus morhua"
  )
```

## References

Froese, R. and D. Pauly (eds.) FishBase. World Wide Web electronic publication. www.fishbase.org.

Robinson, R.A. et al. (2010). Trophic relationships of marine benthic invertebrates in the North Sea. *Journal of the Marine Biological Association of the United Kingdom*, 90(7), 1375-1388.
